EuPPollNet: A European database of plant-pollinator networks

### 2024-06-10 Manuscript/EuPPollNet_Ms.qmd
Edits

---
title: "EuPPollNet: A European database of plant-pollinator networks"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}

bibliography: references.bib
latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---

@@

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising the long-term persistence of plant-pollinator interaction networks. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time, along with other ecological variables of interest. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences. Furthermore, this work assesses the taxonomic and sampling coverage of the database at the European level, and the key structural properties in plant-pollinator networks. We hope this database will facilitate future research that fills taxonomic, ecological, and geographical data gaps on plant-pollinator interactions. Also, we believe it will help evaluate patterns and drivers of plant-pollinator network change and guide future conservation planning for both plant and pollinator species.

@@

**Software format:** The database was built with the R programming language and is stored as “.rds” and “.csv” formats. The construction of the database is fully reproducible and can be accessed at the following PERMANENT LINKS.

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. Hence, this emphasizes the importance of evaluating the spatio-temporal coverage of interactions to guide future sampling efforts and accurately understand the processes that shape plant-pollinator interactions.

@@

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at the European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

JUSTIFY WITH A SENTENCE HERE:
Check that the poll and flower number info is in results! and the phylo signal too.

In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitude range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To keep high quality standards that will allow robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). Each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file (.csv or .rds) that can be merged with the interaction data through the "Flower_data_merger" column. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The potential number of pollinator species at the European level was estimated by adding the total number of species of bees, syrphids, and flies from the checklists, along with the extrapolated number of species from other insect pollinators. This extrapolated number was estimated by assuming that the sampling coverage of these other insect pollinators is equal to the average coverage across bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

@@

The completeness of the EuPPollNet database was evaluated by exploring the rarefied accumulation curves of plant and pollinator species and their interactions across the different sampling sites. In addition, an outstanding question in ecology is how many pollinators are required to pollinate flowering plants [e.g., @kleijn2015 for crops]. To that end, we also calculated the accumulation curve of pollinator species with an increasing number of plant species. The rarefied and extrapolated sampling curves were obtained using the **iNEXT** package [@hsieh2016]. This was complemented with 100 random accumulation curves obtained programmatically to guide the visualization of the different rarefied curves.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness and examined how these network metrics change across the latitudinal range of studies and bioclimatic regions. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. Second, we used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we explored the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

@@

EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). In addition, the presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also seems not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

The estimated sampling coverage of plant and pollinator species within EuPPollNet across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). When we consider the accumulated pollinator richness across sampled plant species, this curve also shows a “quasi-asymptotic” growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). We find that a small portion of plant species and pollinator species are shared across a broad range of sampling sites and that most plant (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinator (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) species are exclusively found in less than 1% of sampling sites (**Figures 4e-4f**). The most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 4}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their interactions across sampling sites. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represent the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrate the extrapolated curve for approximately 3000 sampling sites. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across an increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across sampling sites. Species on the left (i.e., common) are found in many sampling sites, while species on the right (i.e., rare) are found in few or only a single sampling site. Note that indeed \\textit{Apis mellifera} is the most common pollinator but was excluded from this visualization.", fig.width=10, fig.height=6, out.width="115%"}

@@

#Order levels
d_poll1_lines$col = factor(d_poll1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p1 = ggplot(d_poll1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col),size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_poll1_point, col = "black", size=2) + 
ylab("Pollinator species") +
xlab("Sampling sites") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000)+
ylim(0, 3000) +
ggtitle("(a)") +
theme(plot.margin = margin(r = 12))

@@

#Order levels
d_plant1_lines$col = factor(d_plant1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p2 = ggplot(d_plant1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_plant1_point, col = "black", size=2) + 
ylab("Plant species") +
xlab("Sampling sites") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000)+
ylim(0, 3000)+
ggtitle("(b)")+
theme(plot.margin = margin(r = 12))

@@

#Plot
#Order levels
d_int1_lines$col = factor(d_int1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p3 = ggplot(d_int1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_int1_point, col = "black", size=2) + 
ylab("Interactions") +
xlab("Sampling sites") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000) +
ylim(0, 30000)+
ggtitle("(c)")

@@

All data and code to produce of this database and manuscript are available at Zenodo (LINK) and Github (LINK).

@@

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="90%", fig.cap= "\\textbf{Figure S2}. Barplot indicating the number of sampling sites by habitat and bioclimatic region within the database.", fig.width=8, fig.height=8}

@@

ggplot(c, aes(Habitat_bioregion, Count)) +
geom_col(fill = "gray25",width = 0.9, color="black", linewidth = 0.4) +
theme_bw()+
coord_flip(expand = F) +
ggforce::facet_col(~ factor(Group, levels=c("Habitat","Bioregion")),
        space = "free", scales = "free_y") +
ylab("Sampling sites") +
xlab(NULL) +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
      axis.ticks.x = element_line(color="black"),
      strip.text.x = element_text(size=14, face = "bold"),
      strip.background = element_rect(size=0.5, linetype="solid"),
  axis.title = element_text(face = "bold", size = 14),
axis.text = element_text(size=10))

>>>>

---
title: "EuPPollNet: A European database of plant-pollinator networks"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{caption}
        \captionsetup[figure]{labelformat=empty}
        \captionsetup[table]{labelformat=empty}
        \usepackage{lineno}
        \linenumbers
bibliography: references.bib
latex-tinytex: false
link-citations: TRUE
linkcolor: RoyalBlue
urlcolor: RoyalBlue
---

@@

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising the long-term persistence of plant-pollinator interaction networks. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time, along with other ecological variables of interest. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences. We present the taxonomic and sampling coverage of EuPPollNet, and summarize key structural properties in plant-pollinator networks. We hope EuPPollNet will stimulate future research that fills the taxonomic, ecological, and geographical data gaps on plant-pollinator interactions that we have identified. Further, the variation in the structure of the networks in EuPPollNet provides a strong basis for future studies aimed at quantifying drivers of plant-pollinator network change and guiding future conservation planning for plants and pollinators.

@@

**Software format:** The database was built with the R programming language and is stored as “.rds” and “.csv” formats. The construction of the database is fully reproducible and can be accessed at the following link: <https://github.com/JoseBSL/EuPPollNet.git>.

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches make significant contributions to knowledge, such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

@@

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

We investigated how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across latitudes and bioclimatic regions. Networks that are more nested than expected by chance exhibit greater redundancy in their ecological functions and more stability in the face of species loss. EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of nestedness, which has only been evaluated with a relatively small number of networks [@bascompte2003; @payrato2019]. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To maintain high quality standards that will support robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

The EuPPollNet database is available in both .csv and .rds formats. The files contain a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). Each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file (.csv or .rds) that can be merged with the interaction data through the "Flower_data_merger" column. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The number of pollinator species at the European level was estimated by first estimating coverage for the three taxa that have checklists (bees, syrphids and butterflies). The mean coverage across these three groups (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`) was then used for the other taxa for which there is no checklist to extrapolate the number of pollinating species in Europe.

@@

The completeness of the EuPPollNet database was evaluated by exploring the rarefied accumulation curves of plant and pollinator species and their interactions across the different networks. In addition, an outstanding question in ecology is how many pollinators are required to pollinate flowering plants [e.g., @kleijn2015 for crops]. To that end, we also calculated the accumulation curve of pollinator species with an increasing number of plant species. The rarefied and extrapolated sampling curves were obtained using the **iNEXT** package [@hsieh2016]. The different rarefied curves were complemented with 100 boostrapped accumulation curves.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network, and examined how these network metrics change across latitude of studies and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we quantified the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

@@

Europe contains approximately 5,000 species of pollinators and 25,000 species of plants that rely on animal pollination. EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). The presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also does not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

The estimated sampling coverage of plant and pollinator species within EuPPollNet across the different networks is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). When we consider the accumulated pollinator richness across sampled plant species, this curve also shows a “quasi-asymptotic” growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). We find that a small portion of plant species and pollinator species are shared across a broad range of networks and that most plant (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinator (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) species are exclusively found in less than 1% of networks (**Figures 4e-4f**). The most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`* when excluding *Apis mellifera*) species are found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of networks, respectively.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 4}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their interactions across networks. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represent the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrate the extrapolated curve for approximately 3000 networks. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across an increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across networks. Species on the left (i.e., common) are found in many networks, while species on the right (i.e., rare) are found in few or only a single network. Note that indeed \\textit{Apis mellifera} is the most common pollinator but was excluded from this visualization.", fig.width=10, fig.height=6, out.width="115%"}

@@

#Order levels
d_poll1_lines$col = factor(d_poll1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p1 = ggplot(d_poll1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col),size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_poll1_point, col = "black", size=2) + 
ylab("Pollinator species") +
xlab("Networks") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000)+
ylim(0, 3000) +
ggtitle("(a)") +
theme(plot.margin = margin(r = 12))

@@

#Order levels
d_plant1_lines$col = factor(d_plant1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p2 = ggplot(d_plant1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_plant1_point, col = "black", size=2) + 
ylab("Plant species") +
xlab("Networks") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000)+
ylim(0, 3000)+
ggtitle("(b)")+
theme(plot.margin = margin(r = 12))

@@

#Plot
#Order levels
d_int1_lines$col = factor(d_int1_lines$col, levels = c("collectors", "interpolated", "extrapolated"))
#Plot
p3 = ggplot(d_int1_lines, aes(x = t, y = qD, group = method, color = col)) +
geom_line(aes(linetype = col), size = 0.65) +
scale_colour_manual(values = c("gray", "black", "red"), labels = c("Randomizations", "Interpolated", "Extrapolated")) +
scale_linetype_manual(values = c("solid", "solid", "dashed"),  labels = c("Randomizations", "Interpolated", "Extrapolated")) + 
geom_ribbon(aes(ymin = qD.LCL, 
    ymax=qD.UCL, fill = col),alpha=0.2, colour = NA, show.legend = FALSE)+
scale_fill_manual(values = c("gray", "black", "red")) +
geom_point(data = d_int1_point, col = "black", size=2) + 
ylab("Interactions") +
xlab("Networks") +
theme_bw() +
coord_cartesian(expand = FALSE, clip = "off")  +
xlim(0,3000) +
ylim(0, 30000)+
ggtitle("(c)")

@@

All data and code to produce of this database and manuscript are available at Zenodo (LINK) and Github (<https://github.com/JoseBSL/EuPPollNet.git>).

@@

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="90%", fig.cap= "\\textbf{Figure S2}. Barplot indicating the number of networks by habitat and bioclimatic region within the database.", fig.width=8, fig.height=8}

@@

ggplot(c, aes(Habitat_bioregion, Count)) +
geom_col(fill = "gray25",width = 0.9, color="black", linewidth = 0.4) +
theme_bw()+
coord_flip(expand = F) +
ggforce::facet_col(~ factor(Group, levels=c("Habitat","Bioregion")),
        space = "free", scales = "free_y") +
ylab("Networks") +
xlab(NULL) +
theme(plot.margin=unit(c(0,0,0,0), "mm"),
      axis.ticks.x = element_line(color="black"),
      strip.text.x = element_text(size=14, face = "bold"),
      strip.background = element_rect(size=0.5, linetype="solid"),
  axis.title = element_text(face = "bold", size = 14),
axis.text = element_text(size=10))
